---
title: "mfinn13_FinalHomeworkCode_01"
output: html_document
date: "2025-02-10"
---

**Vectors:** essentially a list of data which are all of the same data-type

**Matrix:** a collection of vectors stored together, can also only hold one type of data. Two-dimensional. 

```{r}
m <- matrix(data = c(1,2,3,4), nrow=2, ncol=2)
m
```

Matrices naturally put the data by column, but by inputting the argument byrow = TRUE, the data will be organized by row.

Matrices can also be created by binding vectors using rbind(vector, vector2) or cbind(vector, vector2). cbind arranges by column.

```{r}
v1 <- c(1,2,3,4)
v2 <- c(5,6,7,8)
m1 <- rbind(v1, v2)
m1
m2 <-cbind(v1,v2)
m2
```

**Various matrix-related commands**
class()
dim()
names()
rownames()
colnames()

Structure function: str()
Can be used for any data structure to provide details about it.

**Arrays** hold only one type of data. A vector is a 1-dimensional array, while a matrix is a 2-dimensional array. 

Constructing an array: array(data = *data*, dim = c(*number of rows per matrix*, *number of columns per matrix*, and *number of matrices*))

Example:

```{r}
a <- array(data = 1:90, dim = c(5,6,3))
a
```
**Subsetting**
You can select a subset of a data structure using brackets [like this]. 

**Challenge 1**: Extracting every third word from a sentence
```{r}
ulysses <- c("Stately","plump","Buck","Mulligan","came","from","the","stairhead","bearing","a","bowl","of","lather","on","which","a","mirror","and","a","razor","lay","crossed")
#here I am creating the vector with the words from ulysses.

ulysses2 <- ulysses[seq(from = 3, to = 22, by =3)]
#here I am using the sequence function to select every third word from the vector
ulysses2

```
**Challenge 2:** extracting numbers from a matrix. 
```{r}
m <- matrix(data = 1:80, nrow = 8, ncol = 10, byrow = FALSE)
#starting matrix
m

x <- m[,c(2,3,6)]
x
##matrix with the 2nd, 3rd, and 6th columns from m. I originally coded this less efficiently. I coded it similarly to how I coded the 6,7,8 matrix below.
x

x <- c(m[6, ], m[7, ], m[8, ])
##matrix with rows 6, 7, and 8
x

x <- c(m[2,2]:m[6,9])
##matrix from 2,2 to 6,9
x
```
**Challenge 3** 
Creating a 4-dimensional, 400 element array with the numbers 400 to 1 descending. 
```{r}
x <- array(data = 400:1, dim = c(5,5,4,4))
x
#this creates an array with 4 dimensions. It is essentially 4 groups of 4 2-d arrays, with each 2-d array being 5 rows by 5 columns. There are 16 total 2-d arrays.
```

**Overwriting**
Replacing elements and/or series of elements in a data structure. Ex: a[row,col] <- 4

**LISTS**
Lists can group together a mix of R structures and types of data. They can group a vector, a matrix, and a boolean for example.
```{r}
s <- c("this", "is", "a", "vector", "of", "strings")  # here we have a vector
m <- matrix(data = 1:40, nrow = 5, ncol = 8)  # this is a matrix
b <- FALSE  #and this is a logical statement
l <- list(s, m, b)  # this puts them all together as a list
l
```
You can extract elements from a list just like from an array, except with double brackets. You can also access elements with a data structure that is within the list. 
```{r}
l[[2]][2, 6]  #this will show us the entry at row 2, column 6 of the matrix in the second position of the list
```
You can name the elements in a list, and use those names to call those elements, and even elements within them.

**Challenge 4**
Primate Taxonomy using lists
```{r}
Platyrrhini <- c("Cebidae","Atelidae","Pitheciidae")
Catarrhini <- c("Cercopithecidae","Hylobatidae","Hominidae")
Tarsioidea <- c("Tarsiidae")
#first I created the three vectors that are the most basic level

Anthropoidea <- list(Platyrrhini, Catarrhini)
#here, I added the platyrrhini and catarrhini to a new list, anthropoidea


names(Anthropoidea) <- list("Platyrrhini", "Catarrhini")
#I think this just gives the two elements within anthropoidea their appropriate titles, which appears when printed.

Haplorhini <- list(Anthropoidea, Tarsioidea)
names(Haplorhini) <- list("Anthropoidea", "Tarsioidea")
#this is a similar process, adding anthropoidea and tarsioidea to a new list, haplorhini, and giving them titles which will be printed

Lemuroidea <- c("Cheirogaleidae", "Lepilemuridae", "Indriidae", "Lemuridae",
    "Daubentoniidae")
Lorisoidea <- c("Lorisidae", "Galagidae")
Strepsirhini <- list(Lemuroidea, Lorisoidea)
names(Strepsirhini) <- list("Lemuroidea", "Lorisoidea")
Primates <- list(Strepsirhini, Haplorhini)
names(Primates) <- list("Strepsirhini", "Haplorhini")
#this line I added myself ^
Primates
#these 7 lines I just copied from the module

```

**Data Frames**
A very useful tool. Columns have variables and rows have cases/observations. Uses data.frame() command. 

Can be manually created or from an existing external datasheet. 
Individual variables can be called, such as first name or last name or DOB, using the name of the dataframe and ().
Variables and observations can be added using the rbind and cbind functions. 

$ is used to call things from a data structure. Can also be used to add a category or observation to a dataframe. Using Cbind adds a category as a factor while the dollar sign results in adding the category as a character vector. 

**Factors:** numeric codes that R can use internally that correspond to character values. String factors get assigned a unique numeric code. 

You can convert factors to characters and vice versa using as.character() and as.factor()

**Challenge 5**
creating a data frame, and making the second column logical-valued, and th ethird column factor-valued.
```{r}
m3 <- matrix(data = c(3,0,1,23,1,2,33,1,1,42,0,1,41,0,2), nrow = 5, ncol = 3, byrow = TRUE)
m3
df <- data.frame(m3)
df
##data frame constructed, now coerce the second and third columns
##logical vectors are essentially booleans?
as.logical(df[,2])
as.factor(df[,3])
df
str(df)
```

**5 challenges:**

1. The first challenge I encountered was remembering how to actually start a new project/repository and connect it to github. I looked back at module 3 and ended up going with the method where you create it on github first, then use its url to connect it to rstudio.
2. A second challenge I encountered was actually seeing my commits on github. Sometimes when I could click 'push' on the git tab, I wouldn't actually see it on github. Sometimes I have to go into the tools tab above and kind of manually push the commits. 
3. A third challenge I had was figuring out how to make data logical valued. At first, I did not even know what this meant, but I googled it. Once I figured out it was a type of data it seemed logical that there would be an as.logical().
4. I had a hard time with challenge number 5. I'm not sure if I actually changed the type of variables that were actually in the data frame.
5. I at first had trouble in challenge 2 figuring out how to extract columns and rows as opposed to just one value. 
